Why do people punish? Evidence for a range of strategic concerns

### 2023-11-24 manuscript.tex
finished manuscript first draft

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{x\newline\indent Word count: xxxx words}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{array}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[normalem]{ulem}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{calc}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{wrapfig}
\usepackage{adjustbox}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{makecell}
\renewcommand\theadfont{}
\raggedbottom
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1,small}}
\captionsetup[figure]{font={stretch=1,small}}
\nolinenumbers
\note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Mapping the punishment strategy space},
  pdfauthor={Scott Claessens*1, Quentin D. Atkinson1, \& Nichola Raihani1,2},
  pdflang={en-EN},
  pdfkeywords={x},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Mapping the punishment strategy space}
\author{Scott Claessens\textsuperscript{*1}, Quentin D. Atkinson\textsuperscript{1}, \& Nichola Raihani\textsuperscript{1,2}}
\date{}

@@

\abstract{%
x
}

@@

Humans cooperate on a scale unparalleled in the animal kingdom. One mechanism thought to sustain this level of cooperation is costly punishment, whereby individuals harm others at a personal cost\textsuperscript{1}, ostensibly encouraging cooperative behaviour from the target (or bystanders\textsuperscript{2--4}) in the future. Punishment therefore offers a route to maintaining or increasing cooperation by changing the payoff structure of social interactions such that it no longer pays to cheat or exploit social partners\textsuperscript{1,5}.

@@

Nevertheless, it remains unclear whether individuals playing economic games use punishment as a behaviour-change tool to enforce cooperation or as a means to achieve other ends. Some have argued that punishment is primarily used to shape future behaviour, either to deter personal harm\textsuperscript{3,9,23} or to uphold normative standards of cooperative behaviour\textsuperscript{20,21,24--27}. But while the \emph{threat} of punishment can have a cooperation-enhancing effect, the \emph{enactment} of this punishment seldom changes the future behaviour of deviant targets\textsuperscript{15}. This calls into question whether punishment primarily operates as a behaviour-change tool or whether it is used to achieve other goals.

Beyond behaviour-shaping concerns, there are a host of other reasons that people may want to punish in economic games. Punishers might be motivated by a desire for retribution rather than deterrence, punishing in proportion to the amount of harm that was personally caused\textsuperscript{28}. Punishment might be driven by concerns about relative payoffs, such as disadvantageous inequity aversion (i.e., avoiding having less than others\textsuperscript{15,29}) and/or general egalitarian preferences (i.e., wanting all participants to receive the same payoffs\textsuperscript{30}). Such concerns about relative payoffs may be activated when participants earn less than cheats in economic games or when there are income disparities in these settings. People might also use punishment for competitive purposes, seeking advantageous inequity for themselves (i.e., having more than others) and/or improving their relative position\textsuperscript{15}.

Common economic game designs have been unable to tease apart these different motives for punishment because participants who interact with cheats in these games experience both losses \emph{and} lower relative payoffs. The typical 1:3 fee-to-fine ratio of punishment in economic games compounds this issue. With this setup, people can simultaneously use punishment to reciprocate losses, to deter others from cheating, and to reduce or reverse disparities in payoffs between themselves and targets. To add to this complexity, it is evident that people use punishment in seemingly disparate ways: punishing when no behaviour change is possible, such as in one-shot games\textsuperscript{12,29,31,32} or in games where the target never learns about the punishment\textsuperscript{33}; punishing those who did not cheat or who over-contributed to collective action (antisocial punishment\textsuperscript{17,34}); punishing in scenarios where they were not personally harmed (third-party punishment\textsuperscript{35}); and punishing in scenarios where disparities in payoffs did not arise from participants' actions\textsuperscript{30,36,37}.

The general conclusion from this research is that there is no one unifying function of costly punishment in humans. Instead, punishment should be thought of as a flexible behavioural tool that serves a variety of functions that are not mutually exclusive\textsuperscript{15}. Due to its multipurpose nature, we should therefore expect variation in punishment strategies in the population, much like the observed variation in social learning strategies\textsuperscript{38}. Some individuals may use punishment as a behaviour shaping tool, for example, while others may use it to reduce or reverse payoff differentials.

This insight raises several underexplored questions. First, which punishment strategies are more frequent in human populations? Second, what traits predict adherence to a particular strategy? Previous work has reported that personality is related to cooperative behaviour\textsuperscript{39} and demographics, political ideology, and religiosity are related to punitive behaviour\textsuperscript{40}, but no research has related these variables to specific punishment strategies. Third, do people have insight into their own punishment strategy? Previous work has argued that people are often unaware of the underlying function of their punitive behaviour, yet they feel compelled to enact it anyway\textsuperscript{28,41}.

Here, we aim to delineate the many possible punishment strategies by asking whether people punish in a manner consistent with a specific strategy and, if so, what other characteristics (personality, social preferences, political orientation) predict the use of different punishment strategies. Table \ref{tab:tableStrategies} summarises the potential functions for costly punishment in economic games and the behavioural strategies they predict. Note that we do not include reputational functions of punishment in this table, such as signalling trustworthiness\textsuperscript{4,42--45}, because our focus is on punishment strategies in anonymous economic games without reputational incentives (but see ref\textsuperscript{46}).

@@

\caption{\label{tab:tableStrategies}xxx}\\

@@

 &  & \multicolumn{2}{c}{\thead{No DI 1 \\ 70-10}} & \multicolumn{2}{c}{\thead{No DI 2 \\ 70-30}} & \multicolumn{2}{c}{\thead{No DI 3 \\ (Computer) \\ 70-30}} & \multicolumn{2}{c}{\thead{No DI 4 \\ (1:1 Fee-Fine) \\ 70-30}} & \multicolumn{2}{c}{\thead{DI \\ 70-50}} & \multicolumn{2}{c}{\thead{3PP \\ 70-70 [100]}} \\

@@

Function & \multicolumn{1}{c}{Behavioural strategy} & \multicolumn{1}{c}{\thead{Take \\ 50-30}} & \multicolumn{1}{c}{\thead{No take \\ 70-10}} & \multicolumn{1}{c}{\thead{Take \\ 50-50}} & \multicolumn{1}{c}{\thead{No take \\ 70-30}} & \multicolumn{1}{c}{\thead{Take \\ 50-50}} & \multicolumn{1}{c}{\thead{No take \\ 70-30}} & \multicolumn{1}{c}{\thead{Take \\ 50-50}} & \multicolumn{1}{c}{\thead{No take \\ 70-30}} & \multicolumn{1}{c}{\thead{Take \\ 50-70}} & \multicolumn{1}{c}{\thead{No take \\ 70-50}} & \multicolumn{1}{c}{\thead{Take \\ 50-90 [100]}} & \multicolumn{1}{c}{\thead{No take \\ 70-70 [100]}}\\

@@

Building on previous designs\textsuperscript{29,31,47,48}, we employ a suite of one-shot economic games where individuals are given the opportunity to punish targets at a personal cost. We carefully designed this suite of games to tease apart the proposed punishment strategies in Table 1, such that each strategy predicts a different pattern of behaviour across all the games. We use the resulting behavioural patterns to discern which punishment strategy participants are employing. We then combine these behavioural patterns with data on demographics, personality, political ideology, religiosity, and self-reported strategy usage.

@@

x

@@

x

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{punishment; cooperation; economic games\newline\indent Word count: 5185 words}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{array}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[normalem]{ulem}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{calc}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{wrapfig}
\usepackage{adjustbox}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{makecell}
\raggedbottom
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1,small}}
\captionsetup[figure]{font={stretch=1,small}}
\nolinenumbers
\note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Why do people punish? Evidence for a range of strategic concerns},
  pdfauthor={Scott Claessens*1, Quentin D. Atkinson1, \& Nichola Raihani1,2},
  pdflang={en-EN},
  pdfkeywords={punishment; cooperation; economic games},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Why do people punish? Evidence for a range of strategic concerns}
\author{Scott Claessens\textsuperscript{*1}, Quentin D. Atkinson\textsuperscript{1}, \& Nichola Raihani\textsuperscript{1,2}}
\date{}

@@

\abstract{%
Costly punishment is thought to be one of the key mechanisms sustaining cooperation in humans. However, the motives for punitive behaviour remain unclear. Punishment is often assumed to be motivated by a desire to convert free riders into cooperators, but it is also consistent with a host of other functions, such as levelling payoffs or increasing one's relative position. We used a suite of six economic games to tease apart the different motives for punishment. Across representative samples from the United Kingdom and the United States, we estimated the frequency of different punishment strategies in the population, finding that egalitarian motives for punishment are more common than behaviour-change motives. Moreover, different punishment strategies were differentially predicted by personality, social preferences, political ideology, and religious views. Self-reports of behaviour in the games suggested that people have some degree of insight into their punishment strategy. These findings highlight the multipurpose nature of human punishment.
}

@@

Humans cooperate on a scale that is unparalleled in the animal kingdom. One
mechanism thought to sustain this level of cooperation is costly punishment,
whereby individuals harm others at a personal cost\textsuperscript{1},
ostensibly encouraging cooperative behaviour from the target (or bystanders\textsuperscript{2--4}) in the future. Punishment
therefore offers a route to maintaining or increasing cooperation by changing
the payoff structure of social interactions such that it no longer pays to cheat
or exploit social partners\textsuperscript{1,5}.

@@

Nevertheless, it remains unclear whether individuals playing economic games use
punishment as a behaviour-change tool to enforce cooperation or as a means to
achieve other ends. Some have argued that punishment is primarily used to shape
future behaviour, either to deter personal harm\textsuperscript{3,9,23} or to uphold normative standards of cooperative behaviour\textsuperscript{20,21,24--27}.
But while the \emph{threat} of punishment can have a cooperation-enhancing effect,
the \emph{enactment} of this punishment does not consistently deter targets from
cheating in the future\textsuperscript{15}. This calls into question whether
punishment primarily operates as a behaviour-change tool or whether it is used
to achieve other goals.

Beyond behaviour-shaping concerns, there are a host of other reasons why people
may want to punish in economic games. Punishers might be motivated by a desire
for retribution rather than deterrence, punishing in proportion to the amount
of harm that was personally caused\textsuperscript{28}. Punishment might be driven
by concerns about relative payoffs, such as disadvantageous inequity aversion
(i.e., avoiding having less than others\textsuperscript{15,29}) and/or
general egalitarian preferences (i.e., wanting all participants to receive the
same payoffs\textsuperscript{30}). Such concerns about relative payoffs may be activated
when participants earn less than cheats in economic games or when there are
income disparities in these settings. People might also use punishment for
competitive purposes, seeking advantageous inequity for themselves (i.e., having
more than others) and/or improving their relative position\textsuperscript{15}.

Common economic game designs have been unable to tease apart these different
motives for punishment because participants who interact with cheats in these
games experience both losses \emph{and} lower relative payoffs. The typical 1:3
fee-fine ratio of punishment in economic games compounds this issue. With
this setup, people can simultaneously use punishment to reciprocate losses, to
deter others from cheating, and to reduce or reverse disparities in payoffs
between themselves and targets. To add to this complexity, it is evident that
people use punishment in seemingly disparate ways: punishing when no behaviour
change is possible, such as in one-shot games\textsuperscript{12,29,31,32}, on the very last round of repeated games\textsuperscript{33}, or in games where the target never learns about the
punishment\textsuperscript{34}; punishing those who did not cheat or who
over-contributed to collective action (antisocial punishment\textsuperscript{17,35}); punishing in scenarios where they were not
personally harmed (third-party punishment\textsuperscript{36}); and punishing in
scenarios where disparities in payoffs did not arise from participants'
actions\textsuperscript{30,37,38}.

The general conclusion from this research is that there is no one unifying
function of costly punishment in humans. Instead, punishment should be thought
of as a flexible behavioural tool that serves a variety of functions that are
not mutually exclusive\textsuperscript{15}. Due to its multipurpose nature, we should
therefore expect variation in punishment strategies in the population, much like
the observed variation in social learning strategies\textsuperscript{39}. Some
individuals may use punishment as a behaviour shaping tool, for example, while
others may use it to reduce or reverse payoff differentials.

This insight raises several underexplored questions. First, which punishment
strategies are more frequent in human populations? Second, what traits predict
adherence to a particular strategy? Previous work has reported that personality
is related to cooperative behaviour\textsuperscript{40} and demographics, political
ideology, and religiosity are related to punitive behaviour\textsuperscript{41}, but
no research has related these variables to specific punishment strategies.
Third, do people have insight into their own punishment strategy? Previous work
has argued that people are often unaware of the underlying function of their
punitive behaviour, yet they feel compelled to enact it anyway\textsuperscript{28,42}.

Here, we aim to delineate nine possible punishment strategies by asking
whether people punish in a manner consistent with a specific strategy and, if
so, what other characteristics (personality, social preferences, political
orientation) predict the use of different punishment strategies. Table
\ref{tab:tableStrategies} summarises the potential functions for costly
punishment in the economic games that we considered, and the behavioural
strategies they predict. Note that Table \ref{tab:tableStrategies} is not
an exhaustive list of all possible punishment strategies: we do not include
reputational functions of punishment in this table, such as signalling
trustworthiness\textsuperscript{4,43--46},
because our focus is on punishment strategies in anonymous economic games without
reputational incentives (but see ref\textsuperscript{47}).

Building on previous designs\textsuperscript{29,31,48,49}, we employ a suite of one-shot economic games where individuals
are given the opportunity to punish targets at a personal cost (Figure
\ref{fig:plotGames}). In each game, targets either steal from another
individual or do nothing. Representative samples of participants from the United
Kingdom (\emph{n} = 1014) and the
United States (\emph{n} = 996)
completed all six games on the online platform Prolific. We carefully designed
the suite of games to tease apart the proposed punishment strategies in Table
\ref{tab:tableStrategies}, such that each strategy predicts a different pattern
of behaviour across all the games (see Methods for more detail about the six
games). We use the resulting behavioural patterns to discern which punishment
strategy participants are employing. We then combine these behavioural patterns
with data on demographics, personality, social preferences, political ideology,
religiosity, and self-reported strategy usage.

\begin{figure}
\includegraphics[width=1\linewidth]{images/games} \caption{\emph{Visual summary of the six economic games.} In all
games, Player 2 either steals £0.20 from Player 1 (the focal player) or does
nothing. Player 1 is then given the option to punish by paying a certain amount
of money to remove money from Player 2 (this money is destroyed). The six games
are variants on this general setup, creating situations where (A) Player 2 is
still worse off by stealing, (B) Player 2 creates equality by stealing, (C) the
computer ``decides'' whether Player 2 steals, (D) the fee-fine ratio is 1:1, (E)
Player 2 is better off by stealing, and (F) Player 2 steals instead from a
third-party.}\label{fig:plotGames}
\end{figure}

@@

\caption{\label{tab:tableStrategies}\scriptsize Summary of the different functions for
punishment and the behavioural strategies they predict. \emph{Games A-F are
the games employed in the current study (see Methods for more details). In each
of the six games, participants are given the opportunity to punish players who
``steal'' and those who do not, meaning that participants make twelve punishment
decisions in total. Each behavioural strategy implies a unique pattern of
punishment across all decisions. Green ticks reflect decisions to punish, red
crosses reflect decisions to not punish. In column headers, payoffs at the first
stage (above) and the second stage (below) are denoted as P1-P2 (or P2-P3 {[}P1{]}
for Game F) where participants take the role of P1 and P2 is the target of
punishment. AI = advantageous inequity, DI = disadvantageous inequity.}}\\

@@

 &  & \multicolumn{2}{c}{\thead{\scriptsize Game A \\ (AI) \\ 70-10}} & \multicolumn{2}{c}{\thead{\scriptsize Game B \\ (Equal) \\ 70-30}} & \multicolumn{2}{c}{\thead{\scriptsize Game C \\ (Computer) \\ 70-30}} & \multicolumn{2}{c}{\thead{\scriptsize Game D \\ (1:1 Fee-Fine) \\ 70-30}} & \multicolumn{2}{c}{\thead{\scriptsize Game E \\ (DI) \\ 70-50}} & \multicolumn{2}{c}{\thead{\scriptsize Game F \\ (Third-Party) \\ 70-70 [100]}} \\

@@

Function & \multicolumn{1}{c}{Behavioural strategy} & \multicolumn{1}{c}{\thead{\scriptsize Steal \\ 50-30}} & \multicolumn{1}{c}{\thead{\scriptsize No steal \\ 70-10}} & \multicolumn{1}{c}{\thead{\scriptsize Steal \\ 50-50}} & \multicolumn{1}{c}{\thead{\scriptsize No steal \\ 70-30}} & \multicolumn{1}{c}{\thead{\scriptsize Steal \\ 50-50}} & \multicolumn{1}{c}{\thead{\scriptsize No steal \\ 70-30}} & \multicolumn{1}{c}{\thead{\scriptsize Steal \\ 50-50}} & \multicolumn{1}{c}{\thead{\scriptsize No steal \\ 70-30}} & \multicolumn{1}{c}{\thead{\scriptsize Steal \\ 50-70}} & \multicolumn{1}{c}{\thead{\scriptsize No steal \\ 70-50}} & \multicolumn{1}{c}{\thead{\scriptsize Steal \\ 50-90 [100]}} & \multicolumn{1}{c}{\thead{\scriptsize No steal \\ 70-70 [100]}}\\

@@

@@

The overall pattern of punitive behaviour in the six economic games was in line
with previous research and very similar across both countries (Figure
\ref{fig:plotDecisions}). Participants were generally more likely to punish
targets who stole from another individual compared to targets who did not steal
(multilevel logistic regression; \emph{b} =
1.93, standard error =
0.27, \emph{p}
\textless{} .001). Participants were also more likely
to punish when targets' stealing behaviour generated inequalities,
specifically in Games E and F (\emph{b} =
2.42, SE = 0.44,
\emph{p} \textless{} .001).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotDecisions-1.pdf}
\caption{\label{fig:plotDecisions}\emph{Overall pattern of punitive behaviour across all six
economic games, split by country.} AI = advantageous inequity, DI =
disadvantageous inequity.}
\end{figure}

We classified participants into a particular strategy if their behaviour across
all twelve decisions matched our behavioural predictions shown in Table
\ref{tab:tableStrategies} exactly. Table \ref{tab:tableStrategyCounts} shows
the proportion of participants following each strategy, with N/A used to
represent participants who did not fit exactly into any particular strategy type.
Overall, 59\%
of our participants could be classified exactly into one of the strategies. The
most common strategy in both countries was to never punish across any of the
games. The next most common strategies were those that care about
minimising payoff differences (avoid disadvantageous inequity, egalitarian).
Less common were the behaviour-shaping strategies (deterrent, norm-enforcing),
the retributive strategy, and the competitive strategies (seek AI, competitive).
Although participants often punished targets who did not steal
in the six games (Figure \ref{fig:plotDecisions}), no participants followed the
antisocial strategy by exclusively punishing targets who did not steal across
\emph{all} games.

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:tableStrategyCounts}Counts and proportions of participants
following each punishment strategy exactly, split by country. N/A implies that
participants were unable to be classified exactly into any of the punishment
strategies.}

\begin{tabular}{lcccc}
\toprule
 & \multicolumn{2}{c}{\thead{United Kingdom \\ (N = 1014)}} & \multicolumn{2}{c}{\thead{United States \\ (N = 996)}} \\
\cmidrule(r){2-3} \cmidrule(r){4-5}
Strategy & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Prop} & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Prop}\\
\midrule
Deterrent & 9 & 0.009 & 6 & 0.006\\
Norm-enforcing & 8 & 0.008 & 16 & 0.016\\
Retributive & 6 & 0.006 & 5 & 0.005\\
Avoid DI & 67 & 0.066 & 62 & 0.062\\
Egalitarian & 65 & 0.064 & 71 & 0.071\\
Seek AI & 2 & 0.002 & 0 & 0.000\\
Competitive & 3 & 0.003 & 1 & 0.001\\
Antisocial & 0 & 0.000 & 0 & 0.000\\
Never punish & 426 & 0.420 & 447 & 0.449\\
N/A & 428 & 0.422 & 388 & 0.390\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

To further investigate the strategies that participants were following, we
examined the most common patterns of punitive behaviour across all twelve
decisions. Supplementary Table \ref{tab:tablePatterns} shows the proportion of
participants following the 25 most common behavioural patterns, including,
where appropriate, the predetermined strategies from Table
\ref{tab:tableStrategies}. In both countries, a common pattern of behaviour
not captured by any of the strategies was punishing only when the target stole
in the third-party game (Game F). Punishment in this game is consistent with an
egalitarian motive, as stealing produces unequal outcomes, but third-party
punishment here is also consistent with norm-enforcing and competitive motives
(see Table \ref{tab:tableStrategies}). Other common behavioural patterns not
captured by our strategies included punishing whenever the target stole across
all games and always punishing in every game irrespective of the targets'
behaviour.

While it is useful to look at exact patterns of behaviour, participants may not
have implemented their chosen punishment strategy with exact precision.
In reality, strategies may have been implemented probabilistically for each
punishment decision. There is also the possibility of implementation errors,
whereby participants occasionally ``slip up'' and make decisions that are
incongruent with a particular strategy. This may explain why some participants
were unable to be classified exactly into a single punishment strategy.

To deal with this complexity and include all observed data in our frequency
estimates, we fitted a Bayesian latent state model to the data. This model
assumes that the nine strategies in Table \ref{tab:tableStrategies} (plus a
``random choice'' strategy that chooses randomly for each decision) are the only
latent strategies and that these are instantiated into observed behaviour
according to the logic in Table \ref{tab:tableStrategies} with some
probability of implementation error (i.e., an intention to punish is implemented
as non-punishment and vice versa). Averaging over all strategies and
incorporating the possibility of implementation errors, the model estimates the
probability of participants following any particular strategy, given the
observed data.

The posterior estimates from the model are presented in Figure
\ref{fig:plotModel1b}. The posterior probabilities for each strategy did not
differ between the two countries. In both countries, the never punish strategy
had the highest probability, followed by the egalitarian strategy. The
norm-enforcing and seek AI strategies were the next most likely, with higher
posterior estimates than the competitive and antisocial strategies. None of the
other strategies differed in their posterior estimates. The same general pattern
emerged when we analysed the full dataset without pre-registered exclusions
(Supplementary Figure \ref{fig:plotModel1a}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotModel1b-1.pdf}
\caption{\label{fig:plotModel1b}\emph{Posterior estimates of the probabilities of following
different punishment strategies from the Bayesian latent state model.} The model
assumes an implementation error rate of 5\%. Points represent posterior medians,
line ranges represent 50\% and 95\% credible intervals.}
\end{figure}

Next, we explored which traits predicted adherence to different punishment
strategies. To answer this question, we included variables capturing
demographics, personality, social preferences, political views, and religious
views as predictors in our Bayesian latent state model. We included each
variable in a separate model, predicting all ten punishment strategies
(the nine from Table 1, plus the `random choice' strategy) simultaneously.

Demographic variables tended to be unrelated to strategy usage: age and gender
did not predict adherence to a particular punishment strategy (Supplementary
Figures \ref{fig:plotAllDems2} and \ref{fig:plotAllDems1}). In the United
States, the never punish strategy was slightly more common among participants
lower in socio-economic status (median posterior slope =
-0.20, 95\% CI {[}-0.38
-0.02{]}) but this effect was small.

Conversely, personality and social preferences were linked to variation in
punishment strategies. When including the Big-6 personality dimensions and
Social Value Orientation (SVO) in the model, we found associations with the
egalitarian, never punish, and random choice strategies (Figure
\ref{fig:plotAllPers2}). Participants higher in SVO were more likely to follow
the egalitarian and the never punish strategies, while those with lower SVO
scores were more likely to enact the random choice strategy. The personality
dimensions of honesty-humility and openness to experience were both positively
associated with following the never punish strategy, while extraversion
negatively predicted this strategy. The effects were mostly similar across
countries, but occasionally differed: for example, in the United States,
but not in the United Kingdom, honesty-humility was positively associated with
following the egalitarian strategy and negatively associated with following
the random choice strategy. Overall, the same pattern of results emerged when
analysing the full dataset without exclusions (Supplementary Figure
\ref{fig:plotAllPers1}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllPers2-1.pdf}
\caption{\label{fig:plotAllPers2}\emph{Posterior slopes from Bayesian latent state models
including Big-6 personality dimensions and Social Value Orientation.} Each row
represents a separate model. All models assume an implementation error rate of
5\%. Points represent posterior medians, line ranges represent 95\% credible
intervals.}
\end{figure}

Political and religious variables were also associated with punishment strategy
(Figure \ref{fig:plotAllPolRel2}). These effects tended to be more pronounced
in the United States. Controlling for Social Dominance Orientation, American
participants higher in Right Wing Authoritarianism were more likely to follow
the strategies avoiding disadvantageous inequity and seeking advantageous
inequity. Participants who stated that they would like to ``bring those below
them {[}on the socio-economic status ladder{]} up a peg'' were more likely to follow
the egalitarian strategy, while American participants higher in Social Dominance
Orientation, Right Wing Authoritarianism, and believing that God controls events
in the world were less likely to follow the egalitarian strategy. In general,
religious and conservative participants were less likely to follow the never
punish strategy. This general pattern of results was replicated with the full
dataset (Supplementary Figure \ref{fig:plotAllPolRel1}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllPolRel2-1.pdf}
\caption{\label{fig:plotAllPolRel2}\emph{Posterior slopes from Bayesian latent state models
including political ideology, views about social inequality, and religiosity.}
Each row represents a separate model aside from Social Dominance Orientation and
Right Wing Authoritarianism, which control for one another within the same
model. All models assume an implementation error rate of 5\%. Points represent
posterior medians, line ranges represent 95\% credible intervals.}
\end{figure}

Finally, we asked whether participants had insight into their own punishment
strategy. In other words, could participants self-report the strategy that they
were following during the games? To answer this question, we included
participants' responses to post-game questions about their strategy as
predictors in the model. As before, each predictor was included in a separate
model, predicting all ten strategies simultaneously.

In general, we found that self-reported strategy usage was positively associated
with the behavioural strategy that participants employed (see Supplementary
Figures \ref{fig:plotSliders1} and \ref{fig:plotSliders2} for the distribution
of responses to self-report questions). Figure \ref{fig:plotAllSliders2} shows
the relationships between self-report questions and the different punishment
strategies, highlighting the combinations where the question matched the
behavioural strategy. We found positive relationships between the self-report
questions and strategy usage for the norm-enforcing, egalitarian, seek
advantageous inequity, never punish, and random choice strategies. The 95\%
credible intervals for other estimates included zero, though these estimates
often trended in a positive direction. The same pattern of results was found
when analysing the full dataset without exclusions (Supplementary Figure
\ref{fig:plotAllSliders1}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllSliders2-1.pdf}
\caption{\label{fig:plotAllSliders2}\emph{Posterior slopes from models including
self-reported strategy usage.} Each row represents a separate model. All models
assume an implementation error rate of 5\%. Highlighted estimates represent
combinations where the self-report question matched the behavioural strategy.
Points represent posterior medians, line ranges represent 95\% credible
intervals.}
\end{figure}

@@

Using a suite of economic games measuring punishment in different situations,
we have shown that punishment does not serve just one function, but instead is
a flexible tool that can be and is used for different purposes\textsuperscript{15}.
Punishment is more akin to a swiss army knife than a hammer, used by some to
enforce norms of cooperation and by others to reduce or even create inequality
between individuals. We found that people's punishment strategy can, to some
extent, be predicted by individual differences in personality, social
preferences, and political and religious views. Moreover, contrary to the view
that people are often unable to articulate the reasons for their punitive
behaviour\textsuperscript{28}, people seem to have some degree of insight into the
strategy they are using. Despite small differences, these general patterns
replicated in samples from both the United Kingdom and the United States.

Among the punitive strategies, the most common were particularly sensitive to
inequality in payoffs, either from a self-referential perspective (i.e., avoid
disadvantageous inequity) or more generally (i.e., egalitarian). This is in
line with previous studies which have highlighted inequity aversion as an
important motivation for punishment in economic games\textsuperscript{29,30}. Personality and social preference variables mapped
onto these strategies in expected ways. Traits associated with other-regarding
concern, such as SVO and honesty-humility, predicted following the egalitarian
strategy, whereas religious and conservative individuals were less likely to
follow this strategy, especially in the United States. Moreover, participants
following the egalitarian strategy were able to self-report this strategy,
though the same was not true for the avoid disadvantageous inequity strategy.

Behaviour shaping strategies, such as deterrence and norm-enforcement, were less
common than strategies sensitive to inequality in our set of games. This was
reflected both in participants' elicited punishment behaviour (Figure
\ref{fig:plotModel1b}) and in their self-reports of their own strategy
(Supplementary Figures \ref{fig:plotSliders1} and \ref{fig:plotSliders2}).
Regarding the predictors of these strategies, we found that demographic,
personality, political, and religious variables tended to be unrelated to
deterrent and norm-enforcing punishment strategies. We also found that
participants had insight into the norm-enforcing strategy, but not the deterrent
strategy. This finding is in line with previous research showing that people
struggle to accurately report the deterrent motivations for their punitive
behaviour\textsuperscript{28}.

Other punitive strategies were less common in our dataset, but some were more
prevalent than others. For example, participants were more likely to use
punishment to seek advantageous inequity than to exclusively harm those who did
not steal (i.e., antisocial punishment). The existence of the ``seek AI'' strategy
in our dataset supports the claim that punishment can also be used as a tool to
increase one's own relative position\textsuperscript{15}. While generally rare, we
found that this motive for punishment was more common among authoritarian
participants, at least in the United States, potentially providing an
explanation for why peer punishment has been found to be more common among
conservatives in previous work\textsuperscript{41}. Moreover, the fact that no
participants in our sample punished non-stealing across all games suggests that
antisocial punishment does not function to harm cooperators specifically, as has
been previously suggested\textsuperscript{17}. Instead, antisocial punishment appears
to be motivated by improving one's relative position in general, which is in
line with work showing that antisocial punishment disappears with a 1:1 fee-fine
ratio\textsuperscript{35}.

The fact that people use punishment for many different reasons poses problems
for the way that punishment is operationalised in classic behavioural economic
game studies. In these studies, a common assumption is that participants will
punish to change the behaviour of cheats. But in reality, people may be choosing
the punishment option to achieve a variety of different goals. The targets of
punishment in these studies are likely well aware that punishment could be
levied for these different reasons and this knowledge may impact their
responses. For example, if cheating targets interpret punishment as serving a
competitive motive, it may elicit retaliation rather than encourage
cooperation\textsuperscript{9,15,50}. This might help to
explain the mixed findings in the field as to whether punishment actually
motivates cheating targets to cooperate in the future\textsuperscript{15}.

It is striking that the most common strategy in our dataset was to never punish.
This is partly because punishment in these games imposes an economic cost for no
tangible benefit. If the fee-fine ratio had been lower such that it was cheaper
to punish, we may have seen more punishment from participants. Indeed,
72\%
of participants following the never punish strategy positively stated that they
didn't want to pay to reduce anyone's bonus but would have done so if it were
free. But the frequency of the never punish strategy perhaps also reflects a
more general aversion to peer punishment, an aversion that has been highlighted
in both WEIRD (Western, educated, industrialised, rich, and democratic) samples\textsuperscript{51,52} and in small-scale societies\textsuperscript{53}.
One reason that people may be averse to peer punishment is that, due to its
multipurpose nature, it may be interpreted as a competitive challenge by targets
and trigger retaliation\textsuperscript{15}. In situations that lack clear
institutional norms to legitimise punishment, such as our economic games and
some situations in the real world, people might abstain from peer punishment to
avoid such retaliation, regardless of whether retaliation is actually possible.
By contrast, institutionalised punishment in small-scale societies often
functions to compensate victims adequately while limiting the potential for
feuds and cycles of retaliation\textsuperscript{54,55}. Future research
should uncover whether people are more willing to punish in these
conventionalised contexts.
